Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-04-20 article/ms.tex
maath

The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
possible variance that could be captured by any single component. In theory
this approaches zero, but in practice it will be set by the minimum pairwise
distance in $D$ dimensions. Specifically, if we consider two data points to be
captured by a component in the mixture, an those two data points are the
closest two data points (in high dimensional space) of any data points, then
the variance captured by that Gaussian component will represent the \emph{minimum}
possible variance that can be captured by any single component.  
In practice we calculate the minimum total pair-wise euclidean distance $d_{min}$
in $D$ dimensions, average it over the dimensions, and approximate the minimum
determinant of the covariance matrix $|\cov_l|$ as
\begin{equation}
    |\cov_{l}| \approx \prod_{d=1}^{D}\frac{d_{min}^2}{4} \quad ,
\end{equation}
\noindent{}which approximately represents the variance between two data points,
but ignores correlations and assumes only a mean distance in $D$ dimensions
(for comuptational cost). With these approximations,
\begin{eqnarray}
    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\frac{d_{min}^2}{4}}  \nonumber \\
    \log{|\cov_{l}|} & \approx & 2D\log{\left(\frac{d_{min}}{2}\right)} \quad ,
\end{eqnarray}
\noindent{}and the upper limit on $I_2$ is:
\begin{eqnarray}
    I_{2} & \leq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\cov_l|} \nonumber \\
    I_{2} & \leq & -DK\left(D+2\right)\log{\left(\frac{d_{min}}{2}\right)} \quad .
\end{eqnarray}

@@

There is no finite upper bound on $I_{3}$ because the negative log likelihood
$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
data. A lower limit, however, does exist.

@@

It can be seen that for some finite mixture, the total message length depends 
on the number of components $K$, the number of data points $N$ and its 
dimensionality $D$, and the parameters of that mixture: $\vec{\weight}$, 
$\vecmean$, and $\veccov$. That is to say that we can re-arrange Equation
\ref{eq:objective_function} to separate our objective function into three
smaller functions, and use information theory to determine limits or 
predictions for those quantities. Specifically,

Re-arranging Equation \ref{eq:objective_function},
we can describe our objective function as being the sum of three functions:
$I_{g}(N, D, K, \vec{\weight})$

given mixture of finite
components, the total message length depends on the

Before we introduce and compare search strategies,
here we will

>>>>

The other extreme on $I_2$ can be approximated by estimating the \emph{minimum}
possible variance that could be captured by any single component. In theory
this approaches zero, but in practice it will be set by the minimum pairwise
distance in $D$ dimensions. Specifically, if we consider two data points to be
captured by a component in the mixture, an those two data points are the
closest two data points (in high dimensional space) of any data points, then
the variance captured by that Gaussian component will represent the \emph{minimum}
possible variance that can be captured by any single component.  
In practice we calculate the minimum total pair-wise euclidean distance $d_{min}$
in $D$ dimensions, average it over the dimensions, and approximate the minimum
determinant of the covariance matrix $|\cov_l|$ as
\begin{equation}
    |\cov_{l}| \approx \prod_{d=1}^{D}\left(\frac{d_{min}}{2}\right)^2 \quad ,
\end{equation}
\noindent{}which approximately represents the variance between two data points,
but ignores correlations and assumes only a mean distance in $D$ dimensions
(for comuptational cost). With these approximations,
\begin{eqnarray}
    \log{|\cov_{l}|} & \approx & \sum_{d=1}^{D}\log{\left(\frac{d_{min}}{2}\right)^2}  \nonumber \\
    \log{|\cov_{l}|} & \approx & 2D\log{\left(\frac{d_{min}}{2}\right)} \quad ,
\end{eqnarray}
\noindent{}and the upper limit on $I_2$ is:
\begin{eqnarray}
    I_{2} & \leq & -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\cov_l|} \nonumber \\
    I_{2} & \leq & -DK\left(D+2\right)\log{\left(\frac{d_{min}}{2}\right)} \quad .
\end{eqnarray}

@@

There is no finite upper bound on $I_{3}$ because the negative log-likelihood
$-\log\likelihood$ will approach infinity for an increasingly poor fit to the
data. A useful lower limit, however, does exist. The negative log-likelihood
in Eq. \ref{eq:log-likelihood} can be expanded as,
\begin{equation}
    -\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right) = \sum_{n=1}^{N}\log\sum_{k=1}^{K}\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left[-\frac{1}{2}\left(\vec{\data}_n - \vecmean_k\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_n - \vecmean_k\right)\right]
\end{equation}
\noindent{}and we can define
\begin{equation}
\chi^{2}_{nk} = \left(\vec{\data}_{n} - \vecmean_{k}\right)^\intercal{}\vec{\cov}_{k}^{-1}\left(\vec{\data}_{n} - \vecmean_{k}\right) \quad .
\end{equation}
For the sake of placing a lower bound on $I_{3}$, let us assume that each data
point will be `mostly' explained by one of the components in the 
mixture\footnote{We model the data with partial assignment, but here we merely 
seek to place a lower bound on the negative log-likelihood.}, but we do not 
necessarily know which component, or what the parameters of that component are. 
In that situation, we can state that the component that \emph{does} explain
the data will have $\chi_{nk}^2 \sim \textrm{O}\left(D\right)$ and $\exp(-\frac{1}{2}\chi^2) \rightarrow 1$
whereas the other components that do not explain the data will have 
$\chi_{nk}^2 >> D$ such that $\exp(-\frac{1}{2}\chi^2) \rightarrow 0$. With
this simplification we can write an approximation for the negative log-likelihood
that does not require knowledge about which data points are associated to what
component, because the number of times that the expression $\exp(-\frac{1}{2}\chi^2)$
will be non-zero is for the $k$th mixture component will be $N\weight_k$. The
approximate lower bound is:
\begin{eqnarray}
    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -\sum_{k=1}^{K}N\weight_{k}\log\left[\frac{\weight_k}{\sqrt{(2\pi)^{D}|\vec{\cov}_k|}}\exp\left(-\frac{1}{2}{D}\right)\right] \\
    -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) & \gtrsim & -N\sum_{k=1}^{K}\weight_{k}\log{\weight_k} + \frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log{|\vec{\cov}_k|} + \frac{1}{2}ND\left(1 + \log{2\pi}\right) \quad . \nonumber
    \label{eq:logl-lower-bound}
\end{eqnarray}
\noindent{}Like before, we can consider the lower bound on $-N\sum_{k=1}^{K}\weight_{k}\log\weight_k$
for any mixture of $K$ components by setting the weight of $K - 1$ mixtures
to the smallest realistic values, where those mixtures encapsulate only two
data points and $\todo{w_{1} = w_2 = w_{\cdots} = w_{K-1} = 2/N}$:

\begin{eqnarray}
-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& -N\left[(K-1)\left(\frac{1}{N}\log\frac{1}{N}\right) + \left(1 - \frac{K-1}{N}\right)\log\left(1 - \frac{K-1}{N}\right)\right] \nonumber \\
-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& (K - 1)\log{N} + (K - 1 - N)\log\left(1-\frac{K-1}{N}\right) \nonumber \\
-N\sum_{k=1}^{K}\weight_{k}\log\weight_k &=& N\log{N} + (K - 1 - N)\log\left(N-K+1\right) \quad .
\end{eqnarray}

The remaining term in Eq. \ref{eq:logl-lower-bound}, $\frac{1}{2}N\sum_{k=1}^{K}\weight_{k}\log|\vec{\cov}_k|$, requires some  consideration.

@@
